Assessment and implementation of forestry road network projection methods

### 2023-01-09 analysis/paper/paper.Rmd
Writing in progress...

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [bodner, dietze, perfict]. Recent efforts to forecast impacts of climate change and anthropogenic disturbance on Canadian forests and forest-dwelling wildlife [@tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_;@michelletti_;recent quebec examples] build on a rich history of spatial-stochastic models of forest growth and succession following disturbance [landis;overview;stms;seles]. Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales are not available.

Forest roads are also essential for harvest planning. Some forest management decision support systems include forest road projection methods [patchworks], but methods implemented in proprietary closed-source commercial software cannot be used in other contexts. Published methods intended to support strategic planning [@Anderson_projecting_2004;Clark et al., 2000; Dean, 1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Clement [2021], and a least cost path minimum spanning tree algorithm used to project roads and reconstruct development history in British Columbia [lochhead published and not]. The former method is available as QGIS and Landis plugins, and the latter is not easily extricated from a larger open source modelling project that is specific to British Columbia []. Neither method has not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].

Comment on metrics relevant to wildlife and what exactly accuracy means.

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development in a mountainous region of British Columbia using performance metrics relevant to wildlife.

#One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).

#Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations. Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes. Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network. In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially.

#value of roadless areas: [@ibisch_a_2016; @talty_conservation_2020; @psaralexi_importance_2017]

#forest mgmt planning models: [@depellegrinllorente_recognizing_2017]

#other of interest: caribou habitat metric [@whitman_a_2017]

#Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area. In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015). Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015).

#Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011).

@@

## Study Area

#TO DO: dynamic links to figs and tables.

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1). This allows us to project road building based on historic forest harvest data. With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) [@brown2010], the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests]. The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population [@brown2010]. The Revelstoke TSA is an example of a landscape where road projections could be used to inform management of critical wildlife habitats.

@@

## Data

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple cost raster based on the slope and major barriers. Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope (Table 1) [@governmentofbritishcolumbia2021a]. Major water bodies were considered barriers and assigned a very high cost. Lastly, existing roads were assigned a cost of 0. The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road construction costs used to build the cost raster.

```{r}
tbl <- "Cost Type|Value ($ per km)|Source
Slope|16178 + 504*slope|Interior appraisal manual
Major lakes|65000|Major water bodies of BC
Existing roads|0|Digital road atlas and forest tenure roads for BC
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df)
```

@@

## Road Projection Methods

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the cost of traversing the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)). Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the road building cost between the two. The least cost path to each cutblock is determined independently of the other cutblocks. The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. The sequence is determined by a heuristic such as the distance to the existing road. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed (Figure \@ref(fig:fig2)). In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method. This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks. Finally, in some cases the impact of forestry on wildlife is projected without considering the road development that will accompany it, we refer to this as the cutblocks only projection method and compare it to projections that do consider roads.

@@

```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks. The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP). Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density. Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the 100 m resolution cost raster by the grey shading."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

## Analysis

We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods. We also explored the effects of several variations of the input parameters by repeating the MST projections for each variant. We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced points or random points. Regular and random  point densities of 1 and 10 pts/km^2^ were tested (Figure \@ref(fig:fig2)). The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings. For the Hardy QGIS plugin we used the 100 m resolution cost raster. We kept all the default settings except for the skidding distance which was set to 200m. Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^. This resulted in a total of 13 road projection variants to be compared (Table 2).

Table 2: Road projection method variations 
```{r}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST|Regular|1|100 m and 1000 m
MST|Regular|10|100 m and 1000 m
MST|Random|1|100 m and 1000 m
MST|Random|10|100 m and 1000 m
MST|Centroid|N/A|100 m and 1000 m
DLCP|Regular|10|100 m 
Hardy QGIS|N/A|~9|100 m
Cutblocks only|N/A|N/A|100 m

In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road. 4) Road disturbance footprint, a 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks. The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes]. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004]. Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package. Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable with out roads. All road characteristics were stored in raster format.

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and at the cutblock level, where road network characteristic rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.

The projections were assessed for their spatial performance in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Secondly, the performance of each projection for the three binary characteristics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed. All pixels in the projected rasters fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed. These categories were summarized for each projection and characteristic and used to calculate the performance metrics sensitivity, precision and F-measure [@lee2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarly or better across the whole landscape. Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

When comparing between projection method variants we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods. The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results. Hardy's QGIS plugin was closest to the observed mean for the road presence and density characteristics but was among the furthest from the observed mean for the distance to the nearest road and road disturbance footprint characteristic. The cutblocks only method showed that not including road development underestimates anthropogenic disturbance.

@@

```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection method variations with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the characteristic and negative values indicate an underestimation. The cublocks only method is not defined for the road density, road presence or distance to road characteristics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

Using difference maps to highlight areas of poor projections we found several potential causes of discrepancy. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)a). However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley. There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)b). In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)c).

@@

As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`). The cutblocks only method performed worst for all the road network characteristics (Figure \@ref(fig:fig5)). For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the F-measure was similar across projection method variants, while sensitivity was higher at high sampling densities and precision was higher at low sampling densities. The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

**NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion **

@@

## What David had
Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

>>>>

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [bodner, dietze, perfict]. Recent efforts to forecast impacts of climate change and anthropogenic disturbance on Canadian forests and forest-dwelling wildlife [@tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_;@michelletti_;recent quebec examples] build on a rich history of spatial-stochastic models of forest growth and succession following disturbance [landis;overview;stms;seles]. Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales have not been available.

Forest roads are also essential for harvest planning. Some forest management decision support systems include forest road projection methods [patchworks], but methods implemented in proprietary closed-source commercial software cannot be used in other contexts. Published methods intended to support strategic planning [@Anderson_projecting_2004;Clark et al., 2000; Dean, 1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Clement [2021], and a least cost path minimum spanning tree algorithm used to project roads and reconstruct development history in British Columbia [lochhead published and not]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia []. Neither method has been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].

<!-- Comment on metrics relevant to wildlife and what exactly accuracy means. -->
In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife.

<!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). -->

<!-- Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations. Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes. Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network. In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially. -->

<!-- value of roadless areas: [@ibisch_a_2016; @talty_conservation_2020; @psaralexi_importance_2017] -->

<!-- forest mgmt planning models: [@depellegrinllorente_recognizing_2017] -->

<!-- other of interest: caribou habitat metric [@whitman_a_2017] -->

<!-- Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area. In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015). Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015). -->

<!-- Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011). -->

@@

## Study Area and Data

<!-- TO DO: dynamic links to figs and tables. -->
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@brown2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests]. The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).

@@

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. All roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@governmentofbritishcolumbia2021a], and that road construction across major water bodies is prohibitively expensive (Table 1). Existing roads are assigned a cost of 0. The cost raster consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. A simple model of road construction costs.
```{r}
tbl <- "Cost Type|Value ($ per km)|Source
Slope|16178 + 504*slope|Interior appraisal manual
Major lakes|65000|Major water bodies of BC
Existing roads|0|Digital road atlas and forest tenure roads for BC
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df)
```

@@

<!-- TO DO: add links to repositories and explain what data is and is not there, and why -->

## Road Projection Methods and Variants

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)). We use Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock is determined independently in simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)). We compared these methods to an alternative iterative least cost path method developed by Hardy [ref].

<!-- TO DO: cite key R packages used in method implementation -->

The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:fig2)). We used default settings in the Hardy QGIS plugin except for kidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table 2).

Table 2: Road projection method variants. 
```{r}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST|Regular|1|100 m and 1000 m
MST|Regular|10|100 m and 1000 m
MST|Random|1|100 m and 1000 m
MST|Random|10|100 m and 1000 m
MST|Centroid|N/A|100 m and 1000 m
DLCP|Regular|10|100 m 
Hardy QGIS|N/A|~9|100 m
Cutblocks only|N/A|N/A|100 m

To make help make simple road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.

<!-- TO DO: add links to R package documentation & resources -->

@@

```{r fig2, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (100 m resolution) is indicated by grey shading."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

## Performance Metrics

In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch2016a]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley2004; more]. Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads. All road characteristics were stored in raster format.

<!-- TO DO: update the details on getDistFromSource function - where is it now? -->
<!-- TO DO: Does the cutblocks only scenario include the old roads? If no, it should. And if so, it seems like the other metrics could be calculated in the cutblocks only scenario. It would be interesting to have the "no projection" baseline for all the metrics -->

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:fig3)). All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks. The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads. All methods underproject road density, in part because within 100 m pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

<!-- How is the 'cutblocks only' scenario overestimating the overall road disturbance footprint? That doesn't make sense. -->

<!-- Discussion pts
- disturbance footprints and distance to road are easier to project than road density and road presence.
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. 
-->

@@

```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. The cutblocks only method is not defined for the road density, road presence or distance to road characteristics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

Difference maps highlight several potential causes of discrepancy between our projections and the observed road network. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)a). However, the cost raster is not sufficiently details to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys. There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)b). In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)c). All these discrepancies could be addressed with more accurate harvest and cost information.

@@

As expected, none of the projection methods can accurately project road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives. All projection methods yielded more accurate results than the 'cutblock only' scenario, indicating that any of these projection methods are better than none (Figure \@ref(fig:fig5)). The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

<!-- NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion -->

@@

<!--  What David had
Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.
